---
title: "Coding Exercise Tobin"
author: "Jorge Ballestero"
date: "2026"
output:
  pdf_document:
    toc: true
    number_sections: true
    extra_dependencies: ["amsmath", "booktabs", "float"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
```{r libraries}
library(tidyverse)
library(fixest)
library(modelsummary)
library(lubridate)
library(ggplot2)
library(kableExtra)
```

# Data & Cleaning

## Data Cleaning

Several cleaning steps are applied uniformly to both the estimation sample and the prediction rows via a shared function, ensuring no discrepancy between the data used for training and for prediction.

**Variable exclusions.** Nine variables are dropped prior to estimation. `sprinkler_coverage_sf` is excluded due to NAs exceeding 10% of the sample and near-zero variation among non-missing observations. `grantor` and `grantee` are seller and buyer names, and the free-text is irrelevant to the regression. The remaining dropped variables are `unit_no`, `pre_direction_code`, `floor_material_type_code`, `interior_finish_type`, `unit_type`, and `completion_percent`. Once again, they display either near-universal missingness or negligible variation across observations, and therefore carry no predictive content. I considered dropping `roof_construction_type_code`, but there is enough variation to warrant keeping it, and the type of roof could correlate with the type of house, and thus, price. 

**Structural outliers.**  To avoid structural outliers pulling the regression in a spurious direction, Iwinsorize `sale_price`, `improvement_sf`, `total_net_acres`, `no_of_bedroom`, `no_of_bathroom`, `no_of_story`, `total_garage_sf`, and `age_at_sale` at the 1 and 99 percentiles. This allows a cleaner output that restricts the effect of outliers. 

**Timeframe trimming** Before 1983, sales per year are less than 100 (except 104 in 1982). I exclude these on the assumption that 1983 corresponds to the beginning---or development---of suburban residential developmant in Douglas County. I restrict our sample to 1983 onwards. 

```{r load-data}
train <- read.csv("data/TrainingData.csv")
test  <- read.csv("data/TestData.csv")
```
```{r clean}
winsorize <- function(x, low = 0.01, high = 0.99) {
  pmin(pmax(x, quantile(x, low,  na.rm = TRUE)),
            quantile(x, high, na.rm = TRUE))
}

clean_data <- function(df) {
  df %>%
    mutate(
      trans_date = dmy(transaction_date),
      year       = year(trans_date),
      month      = month(trans_date)
    ) %>%
    select(-c(sprinkler_coverage_sf, grantor, grantee,
              unit_no, pre_direction_code,
              floor_material_type_code, interior_finish_type,
              unit_type, completion_percent)) %>%
    mutate(
      age_at_sale = pmax(year - built_year, 0),
      remodeled   = as.integer(remodeled_year > 0)
    ) %>%
    filter(
      property_type_code == "Residential",
      vacant_flag        == "N",
      year >= 1983
    ) %>%
    mutate(
      sale_price      = winsorize(sale_price),
      improvement_sf  = winsorize(improvement_sf),
      total_net_acres = winsorize(total_net_acres),
      no_of_bedroom   = winsorize(no_of_bedroom),
      no_of_bathroom  = winsorize(no_of_bathroom),
      no_of_story     = winsorize(no_of_story),
      total_garage_sf = winsorize(total_garage_sf),
      age_at_sale     = winsorize(age_at_sale)
    )
}

train_clean <- clean_data(train) %>% filter(predict == 0, !is.na(sale_price))
```

# Empirical Strategy

## Hedonic Pricing Framework

As per the instructions' hint, I follow the hedonic pricing approach, decomposing each transaction price into the implicit prices of its constituent characteristics. The estimating equation is:

$$\log(\text{price}_{it}) = \alpha + \beta X_{it} + \gamma Z_{it} + \delta_n + \lambda_t + \varepsilon_{it}$$

where $X_{it}$ contains continuous structural characteristics (square footage, age, garage size, basement area), $Z_{it}$ contains categorical characteristics (quality grade, condition, home style), $\delta_n$ are neighborhood fixed effects, and $\lambda_t$ are year fixed effects. I take the log of price to accommodate the right skew of the price distribution.

## Estimation Sample

To guard against in-sample overfitting, and again following the hint, I hold out December 2011 as a validation set. The model is estimated on 1983 through November 2011, validated on December 2011 (for which prices are observed) and then applied to December 2012.

```{r collapse}
thin_neighborhoods <- train_clean %>%
  filter(!(year == 2011 & month == 12)) %>%
  count(neighborhood_code) %>%
  filter(n < 30) %>%
  pull(neighborhood_code)

train_clean <- train_clean %>%
  mutate(
    neighborhood_code = ifelse(
      neighborhood_code %in% thin_neighborhoods,
      "Other", neighborhood_code
    )
  )
```

## Geographic Fixed Effects

Neighborhood fixed effects are included at the `neighborhood_code` level, yielding 109 distinct areas. Nineteen neighborhoods with fewer than 30 transactions in the estimation sample (representing 0.1% of total observations) are collapsed into a single "Other" category to avoid overfitting on thin cells. The final specification includes 91 neighborhood fixed effects.

## Switching Model
```{r models}
val_set <- train_clean %>%
  filter(year == 2011, month == 12)

train_est <- train_clean %>%
  filter(!(year == 2011 & month == 12)) %>%
  mutate(
    quality_code       = factor(quality_code),
    condition          = factor(condition),
    built_as           = factor(built_as),
    hvac               = factor(hvac),
    roof_material_code = factor(roof_material_code),
    neighborhood_code  = factor(neighborhood_code),
    year               = factor(year)
  )

train_est_m2 <- train_est %>%
  filter(as.numeric(as.character(year)) >= 2008)

cat("Estimation observations (full):      ", nrow(train_est),    "\n")
cat("Estimation observations (2008-2011): ", nrow(train_est_m2), "\n")
cat("Validation observations:             ", nrow(val_set),      "\n")

m1 <- feols(
  log(sale_price) ~ improvement_sf + age_at_sale +
                    no_of_bedroom  + no_of_bathroom,
  data = train_est,
  vcov = "HC1"
)

m3 <- feols(
  log(sale_price) ~ log(improvement_sf) + age_at_sale +
                    no_of_bedroom  + no_of_bathroom +
                    no_of_fireplace + total_garage_sf +
                    total_finished_basement_sf +
                    quality_code + condition + built_as |
                    neighborhood_code + year,
  data = train_est,
  vcov = ~neighborhood_code
)

m4 <- feols(
  log(sale_price) ~ log(improvement_sf) + age_at_sale +
                    no_of_bedroom  + no_of_bathroom +
                    no_of_fireplace + total_garage_sf +
                    total_finished_basement_sf +
                    quality_code + condition + built_as |
                    neighborhood_code + year,
  data = train_est_m2,
  vcov = ~neighborhood_code
)
```
```{r switching-model}
train_est_m2 <- train_est_m2 %>%
  mutate(
    is_luxury = quality_code %in% c("Excellent", "Very Good") &
                improvement_sf > 2500
  )

m_standard <- feols(
  log(sale_price) ~ log(improvement_sf) + age_at_sale +
                    no_of_bedroom  + no_of_bathroom +
                    no_of_fireplace + total_garage_sf +
                    total_finished_basement_sf +
                    quality_code + condition + built_as |
                    neighborhood_code + year,
  data = train_est_m2 %>% filter(!is_luxury),
  vcov = ~neighborhood_code
)

m_luxury <- feols(
  log(sale_price) ~ log(improvement_sf) + age_at_sale +
                    no_of_bedroom  + no_of_bathroom +
                    no_of_fireplace + total_garage_sf +
                    total_finished_basement_sf +
                    quality_code + condition + built_as |
                    neighborhood_code + year,
  data = train_est_m2 %>% filter(is_luxury),
  vcov = ~neighborhood_code
)

cat("Standard model observations:", 
    nrow(train_est_m2 %>% filter(!is_luxury)), "\n")
cat("Luxury model observations:  ", 
    nrow(train_est_m2 %>% filter(is_luxury)),  "\n")
```

# Results

## Model Progression

I estimate three benchmark specifications to illustrate the contribution of each modeling choice. **Model 1** is a naive baseline: size and age only, no fixed effects, estimated on the full sample. **Model 3** adds the full set of structural controls, neighborhood fixed effects, and year fixed effects, also on the full 1983–2011 sample. **Model 4** repeats the Model 3 specification but restricts estimation to 2008–2011, the period most relevant to December 2012 market conditions. The preferred prediction models (`m_standard` and `m_luxury`) are variants of Model 4 estimated on non-luxury and luxury subsamples respectively. This allows to better predict houses in different ranges. 

```{r results}
modelsummary(
  list("Baseline" = m1, 
       "Full 1983-2011" = m3, 
       "Full 2008-2011" = m4),
  stars   = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  output  = "kableExtra"
) %>%
  kable_styling(
    font_size    = 6.5,
    latex_options = "scale_down"
  )
```


## Coefficient Interpretation

The $R^2$ progression suggests that size and age alone explain 9.7% of price variation. Neighborhood and year fixed effects triple this to 36.3%. Restricting to 2008--2011 raises it to 84.6%.

In the preferred specification, a 1% increase in living area yields a 0.515% price increase. Bathrooms add ~2% each; bedrooms are insignificant once size is controlled for. Quality grade carries the largest categorical effect: a home rated Average sells for approximately 70% less than one rated Excellent.

# Predictions

## Prediction Methodology

Predictions are generated using the switching model described in Section 2.4. Residential homes are assigned to `m_standard` or `m_luxury` based on quality and size. The 28 condos and townhouses in the prediction set fall outside the model's estimation scope: they are priced using a size-adjusted fallback derived from 2008--2011 median price per square foot for each property type (\$135/sqft for condos, \$140/sqft for townhouses). Full derivation in Appendix A.

## Predicted Price Distribution

The 191 predicted prices range from \$143,748 to \$1,232,684, with a median of \$310,477 and an interquartile range of \$241,366--\$392,518. The distribution is right-skewed, reflecting the small number of luxury homes in the prediction set.



```{r validate}
val_set_clean <- val_set %>%
  mutate(
    quality_code       = factor(quality_code),
    condition          = factor(condition),
    built_as           = factor(built_as),
    hvac               = factor(hvac),
    roof_material_code = factor(roof_material_code),
    neighborhood_code  = factor(neighborhood_code),
    year               = factor(year),
    is_luxury          = quality_code %in% c("Excellent", "Very Good") &
                         improvement_sf > 2500
  )

val_set_clean$predicted_price <- ifelse(
  val_set_clean$is_luxury,
  exp(predict(m_luxury,   newdata = val_set_clean)),
  exp(predict(m_standard, newdata = val_set_clean))
)

val_set_clean <- val_set_clean %>%
  mutate(
    error     = predicted_price - sale_price,
    abs_error = abs(error),
    pct_error = abs(error) / sale_price * 100
  )

val_set_clean %>%
  summarise(
    n         = n(),
    rmse      = round(sqrt(mean(error^2)), 0),
    mae       = round(mean(abs_error), 0),
    mape      = round(mean(pct_error), 1),
    median_pe = round(median(pct_error), 1)
  )
```
```{r predict}
predict_rows_raw <- train %>%
  filter(predict == 1) %>%
  mutate(
    trans_date  = dmy(transaction_date),
    year        = year(trans_date),
    month       = month(trans_date),
    age_at_sale = pmax(year - built_year, 0),
    remodeled   = as.integer(remodeled_year > 0)
  ) %>%
  select(-c(sprinkler_coverage_sf, grantor, grantee,
            unit_no, pre_direction_code,
            floor_material_type_code, interior_finish_type,
            unit_type, completion_percent)) %>%
  mutate(
    improvement_sf  = winsorize(improvement_sf),
    total_net_acres = winsorize(total_net_acres),
    no_of_bedroom   = winsorize(no_of_bedroom),
    no_of_bathroom  = winsorize(no_of_bathroom),
    no_of_story     = winsorize(no_of_story),
    total_garage_sf = winsorize(total_garage_sf),
    age_at_sale     = winsorize(age_at_sale)
  ) %>%
  mutate(
    neighborhood_code = ifelse(
      neighborhood_code %in% thin_neighborhoods,
      "Other", neighborhood_code
    )
  )

cat("Prediction rows:", nrow(predict_rows_raw), "\n")

predict_residential <- predict_rows_raw %>%
  filter(property_type_code == "Residential") %>%
  mutate(
    quality_code       = factor(quality_code),
    condition          = factor(condition),
    built_as           = factor(built_as),
    hvac               = factor(hvac),
    roof_material_code = factor(roof_material_code),
    neighborhood_code  = factor(neighborhood_code),
    year               = factor(year),
    is_luxury          = quality_code %in% c("Excellent", "Very Good") &
                         improvement_sf > 2500
  )

predict_residential$predicted_price <- ifelse(
  predict_residential$is_luxury,
  exp(predict(m_luxury,   newdata = predict_residential)),
  exp(predict(m_standard, newdata = predict_residential))
)

price_per_sf <- list(
  Condo     = 135,
  Townhouse = 140
)

predict_nonres <- predict_rows_raw %>%
  filter(property_type_code != "Residential") %>%
  mutate(
    predicted_price = case_when(
      property_type_code == "Condo"     ~ improvement_sf * price_per_sf$Condo,
      property_type_code == "Townhouse" ~ improvement_sf * price_per_sf$Townhouse
    )
  )

predictions <- bind_rows(
  predict_residential %>% select(property_id, predicted_price),
  predict_nonres      %>% select(property_id, predicted_price)
) %>% arrange(property_id)

cat("Total predictions:", nrow(predictions), "\n")
cat("NAs:", sum(is.na(predictions$predicted_price)), "\n")
cat("Median predicted price: $",
    round(median(predictions$predicted_price)), "\n")

write.csv(predictions, "output/predictions.csv", row.names = FALSE)
```
```{r predictonplot}
ggplot(predictions, aes(x = predicted_price)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Distribution of Predicted Prices — December 2012",
    x     = "Predicted Price",
    y     = "Count"
  ) +
  theme_minimal()

```

# Evaluation

## Overall Performance

The model achieves a median absolute percentage error of 6.8\% on the 191 December 2012 transactions, evaluated against actual sale prices from the test data. One transaction with a recorded sale price below \$50,000 is flagged as a likely non-arms-length transfer that slipped through the initial filter... and excluded from the primary evaluation, yielding a clean sample of 190. RMSE is \$65,444 and MAE is \$39,024. For context, themedian predicted price is \$310,477. This would suggest that our model's typical absolute error of roughly \$21,000.

## Performance By Price Segment

The model performs best in the \$200,000--\$400,000 range, where the bulk of transactions concentrate and median errors are 6.1\% and 5.0\% respectively. Performance degrades at both tails — the sub-\$200,000 segment consists predominantly of condos and townhouses priced via the fallback method, while homes above \$600,000 reflect the systematic underprediction documented in Section 5.3.

## Where The Model Falls Short

There are 2 main limitations: 

(1) The model systematically underpredicts luxury homes. The transactions above \$800,000 carry a median error of 21.5% and a mean underprediction of \$271,586. Log-linear models compress the upper tail of the price distribution, and the luxury subsample of 1,618 observations is thin relative to the complexity of high-end pricing, where unobserved amenities lik views, lot quality, architectural uniqueness drive substantial value that no transaction-level dataset can fully capture.

(2) Since the traning data excluded non-residential homes, the 28 condos and townhouses are priced using a size adjusted fallback. I assume they are priced differently, so the measure takes an estimated price per sq foot. The fallback approximates value well on average but cannot account for unit-specific characteristics such as floor level, view, or HOA fees capture.


```{r evaluate}
results <- predictions %>%
  left_join(test %>% select(property_id, sale_price),
            by = "property_id") %>%
  mutate(
    error     = predicted_price - sale_price,
    abs_error = abs(error),
    pct_error = abs(error) / sale_price * 100
  )

results %>%
  summarise(
    n         = n(),
    rmse      = round(sqrt(mean(error^2, na.rm = TRUE)), 0),
    mae       = round(mean(abs_error, na.rm = TRUE), 0),
    mape      = round(mean(pct_error, na.rm = TRUE), 1),
    median_pe = round(median(pct_error, na.rm = TRUE), 1)
  )

results %>%
  mutate(
    price_bin = cut(sale_price,
                    breaks = c(0, 200000, 300000, 400000,
                               500000, 600000, 800000, Inf),
                    labels = c("<200k", "200-300k", "300-400k",
                               "400-500k", "500-600k", "600-800k",
                               ">800k"))
  ) %>%
  group_by(price_bin) %>%
  summarise(
    n          = n(),
    mape       = round(mean(pct_error, na.rm = TRUE), 1),
    median_pe  = round(median(pct_error, na.rm = TRUE), 1),
    mean_error = round(mean(error, na.rm = TRUE), 0)
  )

results_clean <- results %>% filter(sale_price > 50000)

results_clean %>%
  summarise(
    n         = n(),
    rmse      = round(sqrt(mean(error^2, na.rm = TRUE)), 0),
    mae       = round(mean(abs_error, na.rm = TRUE), 0),
    mape      = round(mean(pct_error, na.rm = TRUE), 1),
    median_pe = round(median(pct_error, na.rm = TRUE), 1)
  )
```

# Appendix

## A. Non-Residential Fallback Derivation

Condos and townhouses in the prediction set belong to neighborhoods with no single-family residential transactions, making neighborhood fixed effects inapplicable. Iinstead price them using median price per square foot derived from 2008--2011 transactions of the same property type in the training data: \$135/sqft for condos (n=1,816) and \$140/sqft for townhouses (n=1,148). The predicted price for each non-residential unit is then $\hat{p}_i = \text{rate} \times \text{improvement\_sf}_i$.

## B. Luxury Classification Rule

A home is classified as luxury if it satisfies both of the following conditions: (1) `quality_code` is Very Good or Excellent, and (2) `improvement_sf` exceeds 2,500 square feet. This threshold was derived as follows: among homes transacting above \$600,000 in the 2008--2011 estimation window, median square footage is 2,561 and the share rated Very Good or Excellent exceeds 30\%. In the prediction set, 10 of 163 residential homes meet both criteria.

## C. Model Selection Note

An intermediate specification (Model2) was estimated during development. It added quality and style controls to Model 1 but excluded fixed effects, yielding an adjusted $R^2$ of 0.170. It was dropped from the final presentation as it does not contribute to the identification argument: the meaningful jump in explanatory power comes from the addition of neighborhood and year fixed effects in Model 3, not from the quality controls alone.